Sample-Efficient Reinforcement Learning \\ with Implicitly Quantized Representations

### 2024-02-02 paper/main.tex
re ordered figures

Model-based RL (MBRL) can significantly improve the sample efficiency of the model-free methods by reasoning about the world and maximizing the fraction of the information stream utilized for learning by utilizing the observations instead of simply trying to learn a policy or a value function to maximize the return \cite{haRecurrentWorldModels2018}. The world model can be used for planning \cite{allen1983planning, basye1992decision}. A prominent idea has been to optimize the evidence lower bound of observation and reward sequences to learn world models that operate on the latent space of a learned Variational Autoencoder  \citep[VAE,][]{kingmaAutoEncoding2014, igl2018deep}. These models rely on maximizing the conditional observation likelihood $p(o_t | s_t$), which is the reconstruction objective. The latent space of the model can be used for both planning with model-predictive control \cite{rubinstein1997optimization, hafnerLearning2019} or policy learning in the imagination of the world model \cite{hafner2019dream}. \citet{finn2017deep} and \citet{ebert2018visual} show that it is possible to plan with models that have been taught to predict videos directly without a VAE-based latent space. Incorporating the model uncertainty into planning through ensembling or sampling can also have a significant positive impact on the performance of model-based RL agents \cite{deisenrothPILCO2011, chua2018deep, janner2019trust, yu2020mopo, clavera2020model}.

Maximizing the ELBO or video reconstruction error to learn a world model can mislead the model to focus on learning task-irrelevant information, which distorts the optimization landscape \cite{zhang2018natural,zintgraf2021varibad}. When the task is complex, planning with reconstruction-based world models is unreliable \cite{lutter2021learning} and struggles to outperform model-free methods \cite{kostrikov2020image, yarats2021mastering}. The problems are aggravated when the environment is partially observable \cite{morad2023popgym}. Planning can be done through models that learn to predict rewards and values for multi-step rollouts \cite {tamar2016value, silver2017predictron, oh2017value, schrittwieser2020mastering}. However, these methods do not utilize the observations for model learning at all, which can make them inefficient.

@@

TD-MPC \cite{hansenTemporalDifferenceLearning2022} use a consistency loss to learn representations for planning with Model Predictive Path Integral together with reward and value functions learned through temporal difference methods \cite{williams2015model}. TD-MPC2 \cite{hansen2023td} presents a series of improvements to TD-MPC to make the world model show scaling benefits and achieve state-of-the-art performance in a multi-task setting. They train world models with up to 317 million parameters and show monotone improvement in the normalized score across multiple environments and model sizes. The TD-MPC family of methods is applicable to real-world robotics \cite{lancaster2023modem} and learn representations that are suitable for transfer learning after adaptation \cite{yang2023movie} or fine-tuning \cite{feng2023finetuning}. \citet{zhu2023repo} demonstrate that the observation encoder can be made more resilient to spurious correlations by restricting the information flow from the observations to the latent representation, and the encoder can be easily aligned at test time in a reward-free manner. The loss function of TD-MPC can also be used to learn a manager in a hierarchical RL setting \cite{chitnis2023iql}.

@@

The plots in \cref{fig:multi-task-pretraining} show that when the encoder and the dynamics function have been pre-trained, the agent learns quicker than when it is trained from scratch. This effect is particularly visible in Quadruped Run and Humanoid Run. In contrast, when the reward term has been included in the pre-training objective, making the representation task-specific, the pre-training proves to be harmful on Quadruped Run. These results indicate that \our's task-agnostic representation is beneficial in continual and lifelong learning settings.
% However, given the breadth of

@@

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\hat{o}_t &= g_\xi(z_t), \\
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} shows that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the overall performance, thus lending further support to the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.

>>>>

Model-based RL (MBRL) can significantly improve the sample efficiency of the model-free methods by reasoning about the world and maximizing the fraction of the information stream utilized for learning by utilizing the observations instead of simply trying to learn a policy or a value function to maximize the return \cite{haRecurrentWorldModels2018}. The world model can be used for planning \cite{allen1983planning, basye1992decision}. A prominent idea has been to optimize the evidence lower bound of observation and reward sequences to learn world models that operate on the latent space of a learned Variational Autoencoder  \citep[VAE,][]{kingmaAutoEncoding2014, igl2018deep}. These models rely on maximizing the conditional observation likelihood $p(o_t | s_t$), which is the reconstruction objective. The latent space of the model can be used for both planning with model-predictive control \cite{rubinstein1997optimization, hafnerLearning2019} or policy learning in the imagination of the world model \cite{hafner2019dream}. \citet{finn2017deep} and \citet{ebert2018visual} show that it is possible to plan with models that have been taught to predict videos directly without a VAE-based latent space.
% Incorporating the model uncertainty into planning through ensembling or sampling can also have a significant positive impact on the performance of model-based RL agents \cite{deisenrothPILCO2011, chua2018deep, janner2019trust, yu2020mopo, clavera2020model}.

Maximizing the ELBO or video reconstruction error to learn a world model can mislead the model to focus on learning task-irrelevant information, which distorts the optimization landscape \cite{zhang2018natural,zintgraf2021varibad}. When the task is complex, planning with reconstruction-based world models is unreliable \cite{lutter2021learning} and struggles to outperform model-free methods \cite{kostrikov2020image, yarats2021mastering}. The problems are aggravated when the environment is partially observable \cite{morad2023popgym}. Planning can be done through models that learn to predict rewards and values for multi-step rollouts \cite {tamar2016value, silver2017predictron, oh2017value, schrittwieser2020mastering}. However, these methods do not utilize the observations for model learning, making them inefficient.

@@

TD-MPC \cite{hansenTemporalDifferenceLearning2022} use a consistency loss to learn representations for planning with Model Predictive Path Integral control together with reward and value functions learned through temporal difference methods \cite{williams2015model}. TD-MPC2 \cite{hansen2023td} presents a series of improvements to TD-MPC to make the world model show scaling benefits and achieve state-of-the-art performance in a multi-task setting. They train world models with up to 317 million parameters and show monotone improvement in the normalized score across multiple environments and model sizes. The TD-MPC family of methods is applicable to real-world robotics \cite{lancaster2023modem} and learn representations that are suitable for transfer learning after adaptation \cite{yang2023movie} or fine-tuning \cite{feng2023finetuning}. \citet{zhu2023repo} demonstrate that the observation encoder can be made more resilient to spurious correlations by restricting the information flow from the observations to the latent representation, and the encoder can be easily aligned at test time in a reward-free manner. The loss function of TD-MPC can also be used to learn a manager in a hierarchical RL setting \cite{chitnis2023iql}.

@@

The plots in \cref{fig:multi-task-pretraining} show that when the encoder and the dynamics function have been pre-trained, the agent learns quicker than when it is trained from scratch. This effect is particularly visible in Quadruped Run and Humanoid Run. In contrast, when the reward term has been included in the pre-training objective, making the representation task-specific, the pre-training proves to be harmful on Quadruped Run. This indicates that \our's task-agnostic representation is beneficial in continual and lifelong learning settings.
% However, given the breadth of

@@

To provide a thorough analysis of \our, we include results where we add a reconstruction term to our representation loss \cref{eq:rep-loss}:
\begin{align*}
	\hat{o}_t &= g_\xi(z_t), \\
	\mathcal{L}_o &= \mathbb{E}_{o_t\sim\mathcal{D}} [\| \hat{o}_t - o_t \|_2^2],
\end{align*}
where $g_\xi$ is a learned observation decoder that takes the latent state as the input and outputs the reconstructed observation. The decoder $g_\xi$ is a standard MLP. We perform reconstruction at each time step in the horizon.
The results in \cref{fig:reconstruction-loss-ablation} shows that in no environments does reconstruction aid learning, and in some environments, such as the difficult Dog Walk and Quadruped Run, including the reconstruction term has a significant detrimental effect on the performance.
This supports the observations of \citet{zhaoSimplifiedTemporalConsistency2023} and \citet{hansenTemporalDifferenceLearning2022} about the lack of need for a reconstruction target in continuous control tasks.
%Moreover, in the quadruped-run and dog-walk environments incorporating a reconstruction term has a detrimental impact on
%overall performance.
